10 years of school closures and consolidations: the impact on school accessibility

### 2024-12-12 paper/Manuscript.tex
draft revised manuscript - for other author reviews

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={School closures and consolidations: the active travel and emission implications of reduced accessibility},
  pdfauthor={Anastasia Soukhov; Christopher D. Higgins; Antonio Páez; Moataz Mohamed},
  pdfkeywords={accessibility; spatial availability; carbon emissions;
journey-to-school; walkability; active transportation; school
consolidation policy; service provision thresholds},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[School closures and consolidations: the active travel and
emission implications of reduced accessibility]{School closures and
consolidations: the active travel and emission implications of reduced
accessibility}

@@

\abstract{Reducing emissions and increasing active travel are top
priorities for many communities. Often, though, these priorities are
confronted with pressures to reduce the fiscal burden of public
services, including schools. Public schools have been strained to do
more with less, and governments are opting to close under-capacity
schools and expand existing schools to accommodate lost capacity. The
objective of this study is to quantify the travel burdens of school
consolidation/closure policies. We do this through the case of Hamilton,
a mid-sized city in Ontario, Canada. Between the years 2011 and 2015,
7\% of the elementary schools closed in the city. We study how these
closures changed the accessibility landscape for the elementary school
aged population, as well as the active travel and emission implications.
In terms of methods, we use spatial availability, a singly-constrained
multimodal measure of spatial accessibility. Thanks to its proportional
allocation mechanism, spatial availability allows for an interpretable
comparison between `before' and `after' the implementation of school
closure/consolidation policy. Analysis is conducted at the parcel-level
with all school seats being eligible and travel behaviour based on
observed home-to-school patterns for motorized and non-motorized modes.
We report estimates of system-wide impacts that result from the
closures, such as potential growth in GHG emission and reduction of
walking minutes. Findings demonstrate that overall, spatial availability
decreased in the city. Though the school consolidation policy may have
resulted in cost-savings in operation and maintenance of facilities,
students, families, and communities pay the price in reduced
opportunities for physical activity and greater motorized
mobility-related pollution.}

@@

Communities are by nature in flux: economic and demographic shifts,
rural and urban configurations, and political pressures all combine to
produce change. These forces spur the re-organization and consolidation
of public facilities in communities around the world where changes to
service provision may improve the quality of service for some but not
for all (Christiaanse 2020; Rosik, Puławska-Obiedowska, and Goliszek
2021). While change is natural, oftentimes top-down consolidation
decisions are driven by operational cost-savings that do not take into
consideration the full extent of losses in benefits.

From this perspective, school closure policies have been pursued by
governments to optimize per-student operational cost-efficiency (Rong et
al. 2022; Dai et al. 2019) but other implications, such as the impact of
closures on property values (Merrall, Higgins, and Páez 2024), their
effects on the social infrastructure of communities (Butler, Kane, and
Cooligan 2019; Irwin and Seasons 2012), and particularly their travel
implications have often been overlooked by decision makers (Bierbaum,
Karner, and Barajas 2021; J. Lee and Lubienski 2017). This seems a
mistake: journey-to-school trips are tremendously important, and travel
for this purpose is currently the largest contributor of carbon
emissions from personal travel after the journey-to-work (Rong et al.
2022; Pantelaki, Claudia Caspani, and Maggi 2024), where it could be
instead an excellent opportunity for physical activity (Desjardins et
al. 2024). Alas, in many communities globally, schools are closing
(Sageman 2022). For instance, 65\% of comprehensive schools (e.g.~grades
1 through 9) in rural Finland since 1990 closed (Autti and
Hyry-Beihammer 2014), 10\% of elementary schools (kindergarten through
grade 8) in Chicago, USA in 2013 (J. Lee and Lubienski 2017), and 7\% of
primary and lower secondary schools (kindergarten through grade 9) were
closed in Denmark between 2010-2011 (Beuchert et al. 2018). School
closures often increase the distance to school for many students and
thus tend to induce motorized trips (Rong et al. 2022).

We focus on the transportation implications of school closure and
consolidation policies. While some research has considered the ``steps
in reserve'', that is, the potential for active travel hidden behind
motorized trips (Morency, Roorda, and Demers 2009; Morency, Demers, and
Lapierre 2007), school closures potentially make actual steps vanish,
sent to the reserve for the foreseeable future. In exchange for less
physical activity, communities get more motorized travel, with the
concomitant costs and burdens, including emissions. Quantification of
losses in potential active travel and associated growth in carbon
emissions is scarce in the literature, and the focus of this paper. In
addition to its empirical contributions, this work makes use of a novel
multimodal and singly-constrained accessibility measure, called spatial
availability (Soukhov et al. 2023; Soukhov et al. 2024), which we use to
estimate the changes in accessibility and travel after a recent wave of
school closures/consolidations in Hamilton, Ontario.

The results of the research indicate that, overall, school seats became
less spatially available in the city. This change happened unevenly,
with some communities being disproportionately affected. Though the
school consolidation policy pursued by the province and implemented by
the local school board may have resulted in cost-savings in operation
and maintenance of facilities, families, students, and communities now
pay the price in terms of lost opportunities for healthier active
travel, as well as increased emissions from motorized travel.

@@

Hamilton is a mid-size city (\textasciitilde570,000 pop) with rural,
suburban and urban characteristics in the province of Ontario, Canada
(Government of Canada 2022). In 2013, the Hamilton-Wentworth District
School Board (HWDSB), the largest public English school board in the
city, released its Long Term Facilities Master Plan (HWDSB 2013). To
ensure ``equitable, affordable and sustainable learning facilities'',
the Master Plan indicated that 80\% of its elementary schools (attended
by students aged 5 to 13) would be subjected to accommodation
evaluations, including an ``Accommodation Review'', the outcome of which
historically has been school closures/consolidations (HWDSB 2013; Craggs
2013; Seasons 2014b). These decisions, underhandedly supported by the
province, were in part motivated by a search for operational savings
based on projected reductions in student population and underutilized
school capacity (Craggs 2012). Hamilton was particularly affected, as
the HWDSB underwent an unprecedented number of Accommodation Reviews
relative to other school boards (Seasons 2014a).

Though the Accommodation Review guidelines outline a community-focused
process to assess the value that a school provides to students, the
community, and the local economy (Seasons 2014b; Ministry of Education
2006), in practice residents felt that a wholesome breakdown of costs to
repair/maintain schools was not made consistently available (Kleinhuis
2013). Further, others pointed to flawed incentive structures motivating
the HWDSB: ``By closing three schools the HWDSB would have a strong case
with the Ministry of Education to receive full funding for a new
school'' (Kleinhuis 2013). Along with insufficient funding to maintain
schools in a state of good repair (Auditor General of Ontario 2015), the
operational savings as a case for school closures is a long-standing
critique of the school ``funding formula'' established by the
conservative provincial government in 1998 (Mackenzie 2018; Irwin and
Seasons 2012). Within a backdrop of budget cuts, the funding formula
pits the financing of school maintenance and student needs against the
financial compensation of staff and the powers of local school boards,
which depend on the province for the allocation of resources (Mackenzie
2018).

In the wave of school accommodation reviews between 2013 and 2016, 12
elementary schools closed in Hamilton. These closures represented a 7\%
and 4\% decline in elementary school locations and capacity between 2011
(before the school closures) and 2016, despite the elementary-aged
student population in the city slightly increasing between 2011 and
2016( from 58,265 students to 58,865 students).

While accommodation reviews were undertaken to reduce operational costs,
how might have the closure of schools impacted local communities? In
this paper, we hypothesize that the school consolidation policy reduced
accessibility to schools for many students. By extension, this loss in
access and the consequent displacement of active trips to school
increased transportation-related emissions. To test these hypotheses,
the paper examines the following: how does the (1) spatial availability,
(2) loss in walk minutes and (3) associated carbon emissions change for
home-to-school trips after a city closes and consolidates schools?
Further, this work examines (4) what communities, based on low-income
prevalence, experienced the greatest losses as a result of this policy.

Spatial availability, a singly-constrained measure of spatial
accessibility (Soukhov et al. 2023; Soukhov et al. 2024), is used to
quantify accessibility before- and after- school closures, and the
associated impacts for the city of Hamilton and the elementary
school-aged student population as a result of a policy with narrowly
defined benefits. The methods used in this paper can be applied by
decision-makers to evaluate the spatial and transport-related impacts of
public service provision policy. Furthermore, to increase the diffusion
of this analysis, all work is transparent and reproducible, following
best practices in the spatial sciences (Brunsdon and Comber 2021;
Antonio Páez 2021). Accordingly, all associated code and data will be
publicly available within the lead author's
\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
repository}

@@

Our focus is to evaluate the impact of school consolidations between
2011 and 2016 on multimodal school spatial availability and
travel-related GHG emission in Hamilton. To facilitate comparison, two
scenarios are generated. Each scenario reflects the conditions in 2011
and 2016 respectively, assuming the school configurations in each year,
as well as average student population and low-income prevalence of
households, residential parcel locations, and estimated modal travel
times. The difference between the two scenarios is quantified in terms
of accessibility, emissions and equity implications and discussed in
context with the school consolidation policy. Data was retrieved at
different levels of spatial granularity from the following sources.

@@

First, schools and associated school catchments were retrieved from the
City's school boards. In Hamilton, the majority of student-aged
population attends a school in one of the two English public school
boards (FAO 2023). One board is referred to as \emph{public} (i.e.,
Hamilton Wentworth District School Board (HWDSB)) in our work and the
other is referred to as \emph{public-catholic} (Hamilton Wentworth
Catholic District School Board (HWCDSB)). These catchments indicate what
residential property is assigned to what school, by default. Families
can decide to attend schools out of catchment; in fact, 21\%-23\% of
motorized school trips are out of catchment according to the regional
travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016
(Data Management Group 2018). The elementary school locations and
catchments were provided by the HWDSB and HWCDSB for the 2010-2011 and
2015-2016 academic years, and are shown in Figure~\ref{fig-Fig1}.

Formally, elementary schools are defined as schools that provide
instruction to any combination of grades between kindergarten to grade 8
(i.e.~typically children aged 5 to 14). As such, elementary includes
middle schools that only instruct grades 6 to 8 (\emph{MidElem}),
primary schools that only instruct kindergarten to grade 5 or 6
(\emph{JrElem}), and all grade elementary schools (\emph{Elem}) which
instruct all grades from kindergarten to grade 8. The number of `seats'
available in the school beyond or below which it will be over/under-
capacity is calculated by the provincial Ministry of Education; this
value is referred to as the on-the-ground capacity (\emph{OTGC}), and is
assigned to each school. The current/historic OTGCs are not publicly
available (Ontario 2017). However, for schools that underwent the full
Accommodation Review process, OTGCs for a given year could be obtained
from publicly available School Information Profiles. A regression model
was used to estimate the OTGCs for schools that lacked OTGC information.
The independent variables in the model were 1) the school's building
footprint (\emph{F}) in units of (\(m^2\)); 2) the school's shortest
Euclidean distance from the centroid of Hamilton's central business
district area (\emph{DistCBD}) defined in units of (m); and 3)
coefficients associated with a binary variable indicating the type of
school grade instruction (\emph{MidElem}, \emph{JrElem}, \emph{Elem}).
The building footprint from an archived spatial data set (Spatial 2015)
and the 2016 footprints were retrieved from OSM (OpenStreetMap 2021).
Because the age of construction for different schools is not available,
we used each school's distance from the centroid of the Downtown
Business Improvement Area (Hamilton 2014) as a proxy for school age as
the oldest buildings are generally located closer to the city centre
(Merrall 2021). Of note, the models were also used to estimate OTGC of
secondary schools (\emph{Sec}) though not included in the scope of this
work; these public institutions provide instruction for grades 9 to 12
(i.e., children aged 14 to 17).

Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describe
the estimated OTGC for public schools and catholic schools respectively.
School status and OTGC summed by catchment is also visualised in
Figure~\ref{fig-Fig1}. As seen in Table \ref{TabA1-OTGC} in the
Appendix, the coefficient of determination for these models is
\(R^2 = 0.999\) and \(R^2 = 0.998\) for the Public \(OTGC_{Public}\) and
Catholic \(OTGC_{Catholic}\) Boards respectively, and the residual
standard deviation is \(\sigma = 0.212\) and \(sigma = 0.331\).

\begin{equation}
\label{eq:OTGC-catholic}
OTGC_{Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
\end{equation}

@@

\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig1-1.pdf}

@@

\caption{\label{fig-Fig1}The school catchments, estimated On The Ground
Capacity (OTGC) of schools summed per catchment, and school status for
both public and public catholic school boards for year 2011 and 2016.
OTGC scale is presented by quartiles.}

@@

In addition to OTGC for schools, statistics about the student population
and proportion of low-income after tax (LIM-AT) households were
retrieved from the 2011 and 2016 Canadian Census (Canada 2011, 2016)
using the \{cancensus\} \texttt{R} package (Bergmann, Shkolnik, and
Jacobs 2021). The census releases population data by age group category,
so the population aged between 5-9 years and 10-14 years were retained.
LIM-AT refers to the proportion of private households that are below the
median after-tax income (Government of Canada 2017). The LIM-AT
prevalence for households with children under 18 (there is no LIM-AT
prevalence tabulated for households with exclusively elementary aged
children) was retrained. Population and LIM-AT variables were taken at
the Dissemination Area (DA) level, which is the finest geographic unit
publicly available. DAs are designed by Statistics Canada with the aim
of population uniformity hence DAs greatly vary in area but represent
between approximately 400 and 700 (1st and 3rd quartile) in total
population. The population aged 5 to 14 and the proportion of LIM-AT are
visualised in the first and last rows in Figure~\ref{fig-Fig2}.

The third source of data consists of centroids of residential parcels in
2011 and 2016 that are used to represent residential locations (Teranet
2009). There are 134,340 and 139,467 residential locations in 2011 and
2016 respectively, representing 203,806 and 211,596 dwellings (according
to the Canadian Census (Canada 2011, 2016)). As the number of children
per parcel are publicly unavailable, the average number of 5 to 14 year
olds for each DA is divided by the number of residential households in
that DA. Then the potential number of children at each dwelling is
calculated by multiplying this DA rate by the number of dwellings in
each parcel. Due to the proprietary nature of the parcel data, the
middle row in Figure~\ref{fig-Fig2} visualises an aggregation of the
information: the average rate of 5-14 year olds per residential
household at the DA level.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={10 years of school closures and consolidations: the impact on school accessibility},
  pdfauthor={Anastasia Soukhov; Christopher D. Higgins; Antonio Páez; Moataz Mohamed},
  pdfkeywords={accessibility; spatial availability; carbon emissions;
journey-to-school; walkability; active transportation; school
consolidation policy; service provision thresholds},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[10 years of school closures and consolidations: the impact on
school accessibility]{10 years of school closures and consolidations:
the impact on school accessibility}

@@

\abstract{Reducing motorized travel and increasing active travel are top
priorities for many communities. However, these priorities are
confronted with pressures to reduce the fiscal burden of public
services, including schools. Public school boards have been strained to
do more with less, and opting to close under-capacity schools to deliver
educational services more cost-effectively. However, once a school is
closed, students' proximity to schools decreases, and those who
previously lived close enough to walk or cycle can no longer do so. In
this paper, the City of Hamilton, a mid-sized city in Ontario, Canada,
which experienced a 10\% decrease in elementary school locations between
2011 and 2021, serves as a case study. We quanitfiy how these closures
and consolidations have reshaped accessibility for the elementary
school-aged population and discuss their implications for active and
motorized travel. To measure accessibility, spatial availability, a
singly-constrained multimodal measure of spatial accessibility is used.
This method's proportional allocation mechanism enables interpretable
comparison between accessibility `before' and `after' the implementation
of school closure/consolidation policy considering changes in potential
travel time and population. The analysis is conducted at the parcel
level, considering all school seats as eligible and incorporating
observed home-to-school travel patterns for motorized and non-motorized
modes. We present system-wide impact estimates resulting from these
closures. The findings reveal that overall spatial availability in the
city declined. While the school consolidation policy may have achieved
cost savings in facility operation and maintenance, the burden falls on
students, families, and communities through \emph{potentially} reduced
opportunities for physical activity through an active journey to school
and increased motorized travel externalities.}

@@

The amount of services accessible within communities oscillate: economic
and demographic shifts, rural and urban configurations and political
in/action all induce change. These forces spur pressures to
reorganization and consolidation of public facilities that provide
services, potentially improving access for some but not for all
(Christiaanse 2020; Rosik, Puławska-Obiedowska, and Goliszek 2021).
Schools are a compelling case study in this respect. In many communities
globally, school closure and consolidation policies have been pursued by
governments to optimize per-student operational cost-efficiency (Rong et
al. 2022; Dai et al. 2019). Research has highlighted that these polices
have knock-on implications, including the weakening of communities'
social infrastructure (Butler, Kane, and Cooligan 2019; Irwin and
Seasons 2012; Autti and Hyry-Beihammer 2014) and negative
household-level fiscal impacts such as a reduction of property values
(Merrall, Higgins, and Páez 2024). Operationally, savings may be
experienced but they are paid at different levels by students, families,
communities and society.

A school offers many things to a community: but one dimension is the
travel to and from school, often overlooked by decision-makers
considering school closure and consolidation policies (Bierbaum, Karner,
and Barajas 2021; J. Lee and Lubienski 2017). Schools and residential
locations often proximately co-located, the journey-to-school is an
excellent opportunity for physical activity (Desjardins et al. 2024).
However, closures exchange the physical activity of students' realized
active trips or ``reserved'' active trips (Morency, Roorda, and Demers
2009; Morency, Demers, and Lapierre 2007) (e.g., trips that could be
made active) for motorized travel (Rong et al. 2022). Indeed, motorized
travel to school is the largest contributor of carbon emissions from
personal daily travel after the journey-to-work (Rong et al. 2022;
Pantelaki, Claudia Caspani, and Maggi 2024). Despite these societal
implications, many communities are closing and/or consolidating schools
(Sageman 2022). For instance, 65\% of comprehensive schools (e.g.~grades
1 through 9) in rural Finland since 1990 closed (Autti and
Hyry-Beihammer 2014), 10\% of elementary schools (kindergarten through
grade 8) in Chicago, USA in 2013 (J. Lee and Lubienski 2017), and 7\% of
primary and lower secondary schools (kindergarten through grade 9) were
closed in Denmark between 2010-2011 (Beuchert et al. 2018).

In this paper, we explore the case study of Hamilton, a mid-sized city
in Canada that has pursued similar school closure and consolidation
policies. We detail the Canadian political context that incentived these
policies and quantify their transportation implications using data that
constructs where students may have lived and the capacity of schools in
2011, 2016 and 2021. To quantify transportation implications, we use a
novel multimodal extension of a singly-constrained accessibility measure
known as spatial availability (Soukhov et al. 2023; Soukhov et al.
2024). Spatial availability is used to estimate active (i.e., walking)
and motorized (i.e., car) accessibility considering the capacity of each
school, and the residential location and quantity of school-aged
population, and travel times from residential locations to schools for
three time periods. In this way, this work contributes a empirical
findings documenting how government austerity hurts public education,
along with a methodology that results in interpretable accessibility
values that can be compared across modes, the City, and three time
periods.

To foreshadow the paper's conclusions, our findings indicate that school
seats have became less spatially available in the city. This change
happened unevenly, with certain communities being disproportionately
affected. The school closure and consolidation policy pursued may have
achieved cost savings in facility operation and maintenance. However, it
has come at a cost to families, students, and communities, resulting in
lost opportunities for healthier active travel and a potential increase
in motorized travel. We investigate: were the closures and
consolidations worth the operational savings? The methods used in this
paper can be applied by decision-makers to evaluate the spatial and
transport-related impacts of public service provision policy.
Furthermore, to increase the diffusion of this analysis, all work is
transparent and reproducible, following best practices in the spatial
sciences (Brunsdon and Comber 2021; Antonio Páez 2021). Accordingly, all
associated code and data is publicly available within the lead author's
\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
repository}.

@@

\subsection{Overview of proximity and equity in the school accessibility
literature}\label{overview-of-proximity-and-equity-in-the-school-accessibility-literature}

The examination of access to public services is a well established area
of the transportation literature: from parks (El-Murr et al. 2021) to
healthcare (Pereira, Braga, et al. 2021) and associated dimensions of
inequities (Tiznado-Aitken, Munoz, and Hurtubia 2021; Bittencourt and
Giannotti 2023; Kelobonye et al. 2020), spatial accessibility has been
explored in cases studies worldwide.

Accessibility to schools has been a public service explored across
several transportation dimensions. For instance, Marques, Wolf, and
Feitosa (2020) demonstrates that lower-income neighbourhoods, compounded
by other intersecting dimensions of disadvantage, tend to have lower
accessibility to schools compared to more advantaged neighbourhoods in
Portugal. Similar conclusions are drawn in West Virginia, US by Talen
(2001), England by Burgess et al. (2011), Baton Rouge, US by Williams
and Wang (2014), São Paulo and Curitiba, Brazil by Moreno-Monroy,
Lovelace, and Ramos (2018), Pizzol, Giannotti, and Tomasiello (2021),
and Bittencourt and Giannotti (2023), and Santiago, Chile by
Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
demonstrated that when school closures occurs, what schools end up
closing are not a politically neutral decision. For instance, these
spatial trends are explored in the context of England by Pinch (1987),
areas of declining school-aged populations in Dreseden, Germany by
Müller (2011), rural Finnish communities by Autti and Hyry-Beihammer
(2014), and Chicago, US by J. Lee and Lubienski (2017).

The methods applied to measure school accessibility have been varied,
from simple indicators to complex metrics. Beginning from more simple,
Marques, Wolf, and Feitosa (2020) present indicators of school seats per
student along with the shortest distance from origin to schools within a
school catchment, methods also used in Talen (2001) and Burgess et al.
(2011). Authors make refinements to these methods; Müller (2011)
considers travel time by public transit from origins to schools and
Pizzol, Giannotti, and Tomasiello (2021) considers travel times by
different modes and multiple schools from a single origin along with
other indicators related to school quality. By contrast, Williams and
Wang (2014), J. Lee and Lubienski (2017), Moreno-Monroy, Lovelace, and
Ramos (2018), Tiznado-Aitken, Munoz, and Hurtubia (2021), and
Bittencourt and Giannotti (2023) use more complex approaches such as the
two-step floating catchment area (2SFCA). The 2SFCA is popular in
measure in evaluating access to healthcare services (Antonio Páez,
Higgins, and Vivona 2019) and yields a supply-to-demand ratio that
accounts for the population demanding a service and the quantity that is
supplied within a certain distance range (Shen 1998; Luo and Wang 2003).
The consideration of the supply of service is pertinent in the case of
schools and is acknowledged to as a limitation to using simpler
accessibility measures (Pizzol, Giannotti, and Tomasiello 2021).

In addition to competition for school-seats, theoretically, the
advantage students using faster modes (e.g., motorized modes) offer in
terms of a wider range in school-seat choice is also important. Though
some of the reviewed works using 2SFCA approaches consider multiple
modes, explicit attention to the modal \emph{advantage} has not been
discussed. To this end, in this work we use spatial availability, a
singly-constrained measure of spatial accessibility (Soukhov et al.
2023; Soukhov et al. 2024). Spatial availability allows for the
representation of how many `school-seats' are spatially available to
motorized- and active-trip taking elementary school-aged student
population at a spatial-level, facilitating interpretable comparison
across years. Further, spatial availability can be represented as
school-seats per capita, mathematically equivalent to the
supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et al.
(2023)).

\subsection{Hamilton's school closures and
consolidation}\label{hamiltons-school-closures-and-consolidation}

Hamilton is a mid-size city (\textasciitilde570,000 pop) with rural,
suburban and urban characteristics in the province of Ontario, Canada
(Government of Canada 2022a). The city's public English education is
covered by the Hamilton-Wentworth District School Board (HWDSB). In
2013, the HWDSB released its Long Term Facilities Master Plan (HWDSB
2013). To ensure ``equitable, affordable and sustainable learning
facilities'', the Master Plan indicated that 80\% of its elementary
schools (attended by students aged approximately 5 to 14) would be
subject to evaluations, including a ``Pupil Accommodation Review''
(HWDSB 2013). This Master Plan and the slated Pupil Accommodation
Reviews prompted public outcry, as the outcome of Pupil Accommodation
Reviews are typically schools closures and consolidations (Craggs 2013;
Seasons 2014b).

Like all school boards in Ontario, HWDSB relies on funding from the
province which is distributed through a formula based on pupil enrolment
(Mackenzie 2018; Irwin and Seasons 2012). However, it has been
demonstrated that this funding formula has become insufficient to
maintain schools in a state of good repair (Auditor General of Ontario
2015). Further, existing ``top-up'' programs to assist in operation and
maintenance costs (e.g., School Facilities Operation and Renewal Grant)
were entirely phased out by 2018 (TDSB 2024). Within this funding
backdrop, the Pupil Accommodation Review guidelines outlined a
``community-focused'' process to assessing the value that a school
provides to students, the community, and the local economy (Seasons
2014b; Ministry of Education 2006). However in practice, critical
residents felt the outcomes of the Pupil Accommodation Reviews were
predetermined (Thompson, Collins, and Dean 2024): older schools were
slated to be closed, and a wholesome breakdown of costs to repair and
maintain schools were not made available (Kleinhuis 2013). Other critics
pointed to unbalanced incentives motivating the HWDSB: ``By closing
three schools, the HWDSB would have a strong case with the Ministry of
Education to receive full funding for a new school'' (Kleinhuis 2013).
The Ministry's framework promotes the closing and consolidation of
schools rather than cost-effective solutions for keeping older schools
open (Kleinhuis 2013).

The HWDSB and the Hamilton-Wentworth Catholic District School Board
(HWCDSB), which together provide public English education to the
majority of school-aged children in Hamilton (FAO 2023), were
particularly impacted. While Accommodation Reviews were a province-wide
trend during this period, Hamilton's school boards underwent an
unprecedented number of reviews compared to other boards in other
Ontario municipalities. (Seasons 2014a). In the wave of Pupil
Accommodation Reviews between 2013 and 2016, 12 of the 147 elementary
schools closed in Hamilton. Likely in part due to the public
unpopularity of school closures, in 2017, the provincial government
introduced a moratorium on beginning new Pupil Accommodation Reviews. In
the following years, the HWDSB addressed its backlog and closed 12 more
schools between 2017 to 2021 (HWDSB 2023). Though elementary school
locations declined by 10\% between 2011-2021, HWDSB intends to conduct
future Pupil Accommodation Reviews as initially scheduled, waiting until
the end of the moratorium before dates are assigned (HWDSB 2023).

A justification in these Pupil Accommodation Reviews are operational
savings based on ``projected'' reductions in student population and
``under-utilized'' school capacity (Craggs 2012; TDSB 2024; OPSBA 2024).
As later demonstrated in this study, these policies in fact did result
in school capacity per student reducing overall. Indeed, school seats in
the city declined 5\%. It should be noted that though closures dominated
the Pupil Accommodation Review process, some schools were
``consolidated'' e.g., a school or two were closed and another school
was expanded or opened to accommodate students from closed schools.
Taken that school locations decreased more than overall school-seats,
elementary schools in Hamilton are on average larger than they once
were.

In 2024, the moratorium on Pupil Accommodation Reviews is still in place
and school boards are motivated by the same funding conditions. In
absence of new revenue-generating levers and under the same funding
formula with even lower inflation adjusted per-pupil funding, many
school boards are calling for the end of the moratorium (OPSBA 2024;
TDSB 2024; Draaisma 2024). In the meantime, some school boards are
reducing specialized and community-wide programming (TDSB 2024). Under
the status-quo, the end of the moratorium will result in further school
closures. These pertinent decisions, made under an austere funding
formula and flawed school board incentives does not create an
environment where solutions for a sustainable public education system
can thrive.

Though schools have a plethora of benefits not captured by their
maintenance and operational costs (Butler, Kane, and Cooligan 2019;
Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014), this paper's
scope is limited to their travel implications. For students who live in
proximity to schools, their closure or consolidation directly eliminate
the realized or potential active trips that are, or could have been,
made. Closing a school permanently reduces the potential for active
trips and further entrenches reliance on motorized travel. Additionally,
from the perspective of modal equity, school-aged populations with
access to motorized transport gain a speed-time advantage, providing
them with greater flexibility and choice in accessing school-seats.

The spatial and population-level distribution of these impacts are of
question for this paper. Specifically, the paper's aim is to demonstrate
how these policies impacted the motorized and non-motorized (active)
school-seat accessibility landscape. We hypothesize that the school
closure and consolidation policies reduced accessibility to school-seats
for many students. By extension, this loss in access and the
displacement of active trips likely resulted in decreased potential
active travel minutes and increased motorized travel minutes, with these
impacts distributed inequitably across space and populations. To test
these hypotheses, the paper examines the following:

\begin{itemize}
\tightlist
\item
  How has spatial availability changed for home-to-school trips
  following school closures and consolidations?
\item
  What is the extent of the loss in walking minutes for these trips?
\item
  How has motorized travel increased as a result?
\item
  Which communities, particularly households with a high prevalence of
  low-income and hence more likely experiencing transport poverty,
  experienced the greatest losses due to this policy?
\end{itemize}

@@

Our focus is to evaluate the impact of school consolidations between
2011, 2016 and 2021 on multimodal school-seat spatial availability in
Hamilton. To facilitate comparison, the following conditions are
constructed for each year: 1) school configurations, 2) average student
population, low-income prevalence of households, and residential parcel
locations, and 3) estimated multimodal travel behaviour. The following
sub-sections describe these three types of data.

@@

First, elementary schools locations, how the changed throughout the
study period and their capacities are required for each studied year.

Concerning school locations, the City's school boards provided the
authors with locations and associated school catchments for the
2010-2011 and 2015-2016 academic years. Data for 2021 locations was
retrieved from the City's open data portal (Hamilton 2024). In Hamilton,
the majority of student-aged population attend school in one of the two
English public school boards (FAO 2023). One school board is referred as
Hamilton Wentworth District School Board (HWDSB) and the other as
Hamilton Wentworth Catholic District School Board (HWCDSB). School
catchments for the public and public-catholic catchments indicate what
residential property is assigned to what school by default. However,
families can decide to attend schools out of catchment; in fact,
21\%-23\% of motorized school trips for populations aged 5 through 14
are out of catchment according to the regional travel survey (the
Transportation Tomorrow Survey (TTS)) in 2011 and 2016, the most recent
TTS available at the time of writing (Data Management Group 2018).

In addition to the school's location in 2011, 2016 and 2021, all schools
were also tagged with a school status regarding their existence and/or
change in location or school-seat capacity. A summary of the status of
schools through 2011 to 2021 is illustrated in Figure~\ref{fig-Fig1}.
Figure~\ref{fig-Fig1} also displays a count of the number of schools
with the status between years, the number of schools overall present in
each year, and the symbolic representation of a school status that is
used in Figures throughout the paper. Overall, a school can have one of
seven school statuses in any given year. Either: 1) the school did not
change throughout 2011 to 2021, 2) the school expanded (i.e., increased
school-seat capacity) or was 3) relocated and expanded (N.B., schools
expanded only between 2011 through 2016 as they were typically done
\emph{before} school closures and no new Pupil Accommodation Review was
commenced after 2017), 4) the school closed sometime after 2011 and
before 2016 or 5) after 2016 and before 2021, and 6) the school opened
sometime after 2011 and before 2016 or 7) after 2016 and before 2021.

@@

\includegraphics[width=1.2\textwidth,height=\textheight]{../Figures/Fig1.png}

@@

\caption{\label{fig-Fig1}Overview of the number of schools in Hamilton
in 2011, 2016 and 2021 along with the type of school status change..}

@@

The capacity of each school is also required for each simulated year.
This capacity is reflected as the number of `seats' available in the
school. Enrolment exceeding this capacity indicates over-enrolment,
while enrolment below it indicates under-enrolment. This value is
calculated by the provincial Ministry of Education and referred to as
the on-the-ground capacity (OTGC). The current and historic OTGCs for
schools are no longer publicly available (Ontario 2017), so we estimated
them using a multivariate regression model and validated the results
with a set of OTGC values manually gathered. These values were retrieved
for schools that had undergone a completed Pupil Accommodation Review
and had publicly available School Information Profiles.

For schools lacking OTGC values, two regression models were created: one
for public schools and another for public catholic schools. The
independent variables in the model were 1) a dummy variable indicating
the type of school grade instruction (\emph{MidElem}, \emph{JrElem},
\emph{Elem}), 2) the school's building footprint (\emph{F}) in units of
(\(m^2\)), and 3) the school's shortest Euclidean distance from the
centroid of Hamilton's central business district area (\emph{DistCBD})
defined in metres:

\begin{itemize}
\tightlist
\item
  Formally, elementary schools are defined as schools that provide
  instruction to any combination of grades between kindergarten to grade
  8 (i.e.~typically children aged 5 to 14). As such, elementary includes
  middle schools that only instruct grades 6 to 8 (\emph{MidElem}),
  primary schools that only instruct kindergarten to grade 5 or 6
  (\emph{JrElem}), and all grade elementary schools (\emph{Elem}) which
  instruct all grades from kindergarten to grade 8. \emph{MidElem} and
  \emph{JrElem} school grade instruction type is only present in the
  public school board.
\item
  The building footprint \emph{F} was retrieved from an archived spatial
  data set (Spatial 2015) and footprints from newer schools in 2016 and
  2021 from OSM (OpenStreetMap 2021).
\item
  \emph{DistCBD} was calculated `as the crow flies' from each school to
  the centroid of the Hamilton CBD (43.256684°N, 79.869039°W). Notably,
  this variable can also be seen as a proxy for school construction age
  as older buildings in Hamilton are generally located closer to the
  CBD, the ``old'' Hamilton community (Merrall 2021).
\end{itemize}

Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) summarise
the OTGC regression model for public schools and public catholic schools
respectively. As seen in Table \ref{TabA1-OTGC}, the coefficient of
determination for these models are fantastic: \(R^2 = 0.999\) and
\(R^2 = 0.998\) for the public \(OTGC_{Public}\) and public catholic
\(OTGC_{Public Catholic}\) school boards respectively, and the residual
standard deviation is \(\sigma = 0.212\) and \(sigma = 0.331\).

\begin{equation}
\label{eq:OTGC-catholic}
OTGC_{Public Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
\end{equation}

\input{Tab1.tex}
